Place of death in NRS COVID-19 deaths: visualisation for 2023 data

### 2023-07-13 covid_nrs_place_of_death_visualisations_2023.Rmd
reorganised visualisations

# Proportion annual deaths by place

@@

```{r}
(fig_proportion_annual_deaths_by_place <-
  merged_proportions_of_deaths_by_place %>%
  mutate(place_of_death = factor(x = place_of_death, levels = order_of_place_of_death_levels[c(1,3,2,4)])) %>%  # move home to 2nd position to match order in plot
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(linewidth = 1) +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  # theme_minimal() +
  theme(
    legend.position = "top", 
    plot.caption = element_text(size = 10, colour = "gray60")
    ) +
  labs(
    x = "Year", y = "Proportion of yearly deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2023_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)
fig_proportion_annual_deaths_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
```

@@

predicted_data <- tibble(years_since_2015 = seq(from=0, to=30, by = 0.1)) %>%
  mutate(year = 2015 + years_since_2015,
         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015))
         )

@@

actual_and_predicted_data <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
      mutate(data_type = "Historic")
    ,
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
    mutate(
      data_type = "Predicted",
      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
      )
    ,
    merged_proportions_of_deaths_by_place %>%
      select(year, number_of_deaths, place_of_death) %>%
      filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
      mutate(
        number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
        year = 2015 + compute_x_from_y_using_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
        data_type = "Actual",  # note: this is actual figures, but at the extrapolated year
        actual_year = 2020:current_year
      )
  ) %>%
  mutate(data_type = factor(data_type, levels = unique(data_type)))

>>>>

# Number & Proportion annual deaths by place

## Proportion only

@@

```{r}
(fig_proportion_annual_deaths_by_place <-
  merged_proportions_of_deaths_by_place %>%
  filter(place_of_death!="All") %>%
  ggplot(., aes(x=year, y=proportion_of_total, colour=place_of_death)) +
  geom_point() +
  geom_line(linewidth = 1) +
  geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
  scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year))) +
  scale_y_continuous(label = function(x) scales::percent(x = x, accuracy = 1)) +
  scale_colour_viridis_d(option="C") +
  # theme_minimal() +
  theme(
    legend.position = "top", 
    plot.caption = element_text(size = 10, colour = "gray60")
    ) +
  labs(
    x = "Year", y = "Proportion of yearly deaths (%)",
    colour = NULL,
    subtitle = paste0("* ",only_2023_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n")
    ) +
  guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
  NULL
)

fig_proportion_annual_deaths_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_proportion_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 13, height = 10, dpi = 300, units = "cm")
```

## Number & proportion

```{r}
(fig_annual_deaths_by_place <-
  wrap_plots(
    # Number
    merged_proportions_of_deaths_by_place %>%
    ggplot(data = ., 
           aes(x = year, y = number_of_deaths, colour = place_of_death, shape = place_of_death, linetype = place_of_death)
           ) +
    geom_line(linewidth = 1) +
    geom_point(size = rel(2)) +
    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=number_of_deaths, label="*"), inherit.aes = FALSE) +
    theme_minimal() +
    theme(legend.position = "top", legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year)), minor_breaks = NULL) +
    scale_y_continuous(labels = scales::comma, breaks = seq.int(from=0, to=7e4, by=1e4), minor_breaks = seq.int(from=0,t=7e4,by=5e3)) +
    scale_colour_manual(values = c(scales::viridis_pal(option = "C")(4),"grey")) +
    scale_shape_manual(values = c(scales::shape_pal()(4),NA)) +
    scale_linetype_manual(values = c(rep("solid", 4),"dashed")) +
    guides(colour = guide_legend(nrow=2)) +  # spread colour legend over two rows
    labs(
      title = "Number of deaths",
      x = NULL, y = NULL
    ),
    # Proportion
    merged_proportions_of_deaths_by_place %>%
    filter(place_of_death!="All") %>%
    ggplot(data = ., 
           aes(x = year, y = proportion_of_total, colour = place_of_death, shape = place_of_death)
           ) +
    geom_line(linewidth = 1) +
    geom_point(size = rel(2)) +
    geom_text(data = ~filter(.x, year==latest_year), aes(x=latest_year+0.2, y=proportion_of_total, label="*"), inherit.aes = FALSE) +
    theme_minimal() +
    theme(legend.position = "none", legend.title = element_blank(), axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_x_continuous(breaks = sort(unique(merged_proportions_of_deaths_by_place$year)), minor_breaks = NULL) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    scale_colour_viridis_d(option = "C") +
    labs(
      title = "Proportion of deaths",
      x = NULL, y = NULL
    ), 
    guides = "collect"
  ) +
   plot_annotation(theme = theme(legend.position = "top"), subtitle = paste0("* ",only_2023_text),
    caption = paste0(c(death_registrations_text,source_nrs_text),collapse="\n"))
 )

fig_annual_deaths_by_place %>%
save_output_file(filename = file.path(dir_outputs, paste0(c("fig_annual_deaths_by_place", latest_data_modified_date), collapse = "_")), extensions = c(".pdf",".png"), plot = ., width = 20, height = 10, dpi = 300, units = "cm")
```

@@

predicted_data <- tibble(years_since_2015 = seq(from=0, to=30, by = 0.1)) %>%
  mutate(year = 2015 + years_since_2015,
         predicted = predict.lm(object = regression_model, newdata = tibble(years_since_2015), type = "response")
         )

@@

actual_and_predicted_data <-
  bind_rows(
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2015:2019) %>%
      mutate(data_type = "Historic")
    ,
    merged_proportions_of_deaths_by_place %>%
    select(year, number_of_deaths, place_of_death) %>%
    filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
    mutate(
      data_type = "Predicted",
      number_of_deaths = if_else(year %in% 2020:current_year, compute_y_from_x_using_linear_regression(x = year-2015, regression_model = regression_model), number_of_deaths)
      )
    ,
    merged_proportions_of_deaths_by_place %>%
      select(year, number_of_deaths, place_of_death) %>%
      filter(place_of_death == "Home & other non-institution" & year %in% 2020:current_year) %>%
      mutate(
        number_of_deaths = if_else(year == current_year, number_of_deaths / extrapolated_2020_data$proportion_of_deaths_so_far, number_of_deaths),
        year = 2015 + compute_x_from_y_using_linear_regression(y = number_of_deaths, regression_model = regression_model),  # compute predicted X
        data_type = "Actual",  # note: this is actual figures, but at the extrapolated year
        actual_year = 2020:current_year
      )
  ) %>%
  mutate(data_type = factor(data_type, levels = unique(data_type)))
